Neon: Nuclear Norm to Beat Muon

### 2025-05-13 poster/neon_poster.tex
added lost 1/2 to poster
made algorithms smaller to leave more space for plots
New pictures from Nanogpt training

\medskip
\hrule\medskip
\Head{Neon's update rule}\\
Bernstein and Newhouse suggest obtaining the update step for a weight matrix $W$ as a solution to the optimization problem:
\begin{equation}
    \langle G, \delta W \rangle + \frac{\lambda}{2} \norm{\delta W}^2 \to \min_{\delta W}\,,
\end{equation}
where $G$ is a gradient-like matrix obtained via backpropagation. Setting norm to RMS-to-RMS norm (scaled version of Spectral norm) produces Muon. We consider two different choices instead:
\begin{enumerate}
    \item Choosing nuclear norm, $\norm{\cdot}_*$, produces rank-1 update, defined by 
    \begin{equation}\label{eqn:update_star}
        \delta W = -\frac{1}{\lambda} u_1 \sigma_1 v_1^T\,,
    \end{equation}
    where $\sigma_1$ is largest singular value of $G$, and $u_1$, $v_1$ are corresponding singular values.

@@

\noindent\rule[-5pt]{.8\textwidth}{0.4pt}\\
{\bf Algorithm 1} Nuclear-Neon update step for linear layer\\
\noindent\rule[10pt]{.8\textwidth}{0.4pt}
\vspace{-10pt}
\noindent\begin{tabbing}
    {\bf Input:} $\lambda$, gradient-like matrix $G$. \\
    {\bf Output:} Weight matrix update $\delta W$.\\*[\smallskipamount]
    \qquad \= 1.\ $U, \Sigma, V := \textrm{Lanczos}(G, 1)$. \\
    {\bf Return} $-\frac{1}{\lambda}U \Sigma V^T$.
\end{tabbing}
\vspace{-8pt}
\noindent\rule[10pt]{.8\textwidth}{0.4pt}

@@

\noindent\rule[-5pt]{.8\textwidth}{0.4pt}\\
{\bf Algorithm 2} $F*$-Neon update step for linear layer\\
\noindent\rule[10pt]{.8\textwidth}{0.4pt}
\vspace{-10pt}
\noindent\begin{tabbing}
    \qquad\=\qquad\=\qquad\=\kill % Set three tab stops (4em total)
    {\bf Input:} $\lambda$, $r$, gradient-like matrix $G$. \\
    {\bf Output:} Weight matrix update $\delta W$.\\*[\smallskipamount]
    \> 1. $U, \Sigma, V := \textrm{Lanczos}(G, r + 1)$. \\
    \> 2. $s := \sum_{i=1}^{r}\sigma_r$. \\
    \> 3.\ {\bf If} $(r + 1)\sigma_{r + 1} > s${\bf:}\\
    \>\> 4. $r := r + 1$. 5. $\tau := \frac{s + \sigma_{r}}{r + 1}$.\\
    \> 6.\ {\bf Else if} $(r + 1)\sigma_{r - 1} < s${\bf:}\\
    \>\> 7. $r := r - 1$. 8. $\tau := \frac{s - \sigma_{r}}{r + 1}$.\\
    \> 9.\ {\bf Else:}\\
    \>\> 10. $\tau := \frac{s}{r + 1}$\\
    \> 11. $D = [\Sigma - \tau I]_+$. \\*[\smallskipamount]
    {\bf Save} $r$ for the next iteration.\\
    {\bf Return} $-\frac{1}{\lambda}U D V^T$.
\end{tabbing}
\vspace{10pt}
%\noindent\rule[10pt]{.8\textwidth}{0.4pt}

@@

\medskip
\hrule\medskip
\Head{NanoGPT tests}\\
Couple of sentences and a vector(!) picture (reference picture for now).
% \begin{figure}[h]
%     \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/loss_val.png}}
%     \caption{NanoGPT validation loss}
%     \label{fig:nanogpt_loss}
% \end{figure} 
% \begin{figure}[h]
%     \center{\includegraphics[width=0.8\linewidth]{../figs/nanogpt/train_loss.pdf}}
%     \caption{NanoGPT train loss}
%     \label{fig:nanogpt_train_loss}
% \end{figure}
\begin{figure}[h]
    \center{\includegraphics[width=0.8\linewidth]{../figs/nanogpt/val_loss.pdf}}
    \caption{NanoGPT validation loss}
    \label{fig:nanogpt_val_loss}
\end{figure}
\medskip
\hrule\medskip
\Head{Conclusion}\\
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

>>>>

\medskip
\hrule\medskip
\Head{Neon's update rule}\\
Bernstein and Newhouse suggest obtaining the update step for a weight matrix $W$ as a solution to the optimization problem:
\begin{equation}
    \langle G, \delta W \rangle + \frac{\lambda}{2} \norm{\delta W}^2 \to \min_{\delta W}\,,
\end{equation}
where $G$ is a gradient-like matrix obtained via backpropagation. Setting norm to RMS-to-RMS norm (scaled version of Spectral norm) produces Muon. We consider two different choices instead:
\begin{enumerate}
    \item Choosing nuclear norm, $\norm{\cdot}_*$, produces rank-1 update, defined by 
    \begin{equation}\label{eqn:update_star}
        \delta W = -\frac{1}{2\lambda} u_1 \sigma_1 v_1^T\,,
    \end{equation}
    where $\sigma_1$ is largest singular value of $G$, and $u_1$, $v_1$ are corresponding singular values.

@@

{\footnotesize{}
\noindent\rule[-5pt]{.8\textwidth}{0.4pt}\\
{\bf Algorithm 1} Nuclear-Neon update step for linear layer\\
\noindent\rule[5pt]{.8\textwidth}{0.4pt}
\vspace{-15pt}
\noindent\begin{tabbing}
    {\bf Input:} $\lambda$, gradient-like matrix $G$. \\
    {\bf Output:} Weight matrix update $\delta W$.\\*[\smallskipamount]
    \qquad \= 1.\ $U, \Sigma, V := \textrm{Lanczos}(G, 1)$. \\
    {\bf Return} $-\frac{1}{2\lambda}U \Sigma V^T$.
\end{tabbing}
\vspace{-8pt}
\noindent\rule[10pt]{.8\textwidth}{0.4pt}
}

@@

{\footnotesize{}
\noindent\rule[-5pt]{.8\textwidth}{0.4pt}\\
{\bf Algorithm 2} $F*$-Neon update step for linear layer\\
\noindent\rule[5pt]{.8\textwidth}{0.4pt}
\vspace{-15pt}
\noindent\begin{tabbing}
    \qquad\=\qquad\=\qquad\=\kill % Set three tab stops (4em total)
    {\bf Input:} $\lambda$, $r$, gradient-like matrix $G$. \\
    {\bf Output:} Weight matrix update $\delta W$.\\*[\smallskipamount]
    \> 1. $U, \Sigma, V := \textrm{Lanczos}(G, r + 1)$. \\
    \> 2. $s := \sum_{i=1}^{r}\sigma_r$. \\
    \> 3.\ {\bf If} $(r + 1)\sigma_{r + 1} > s${\bf:}\\
    \>\> 4. $r := r + 1$. 5. $\tau := \frac{s + \sigma_{r}}{r + 1}$.\\
    \> 6.\ {\bf Else if} $(r + 1)\sigma_{r - 1} < s${\bf:}\\
    \>\> 7. $r := r - 1$. 8. $\tau := \frac{s - \sigma_{r}}{r + 1}$.\\
    \> 9.\ {\bf Else:}\\
    \>\> 10. $\tau := \frac{s}{r + 1}$\\
    \> 11. $D = [\Sigma - \tau I]_+$. \\*[\smallskipamount]
    {\bf Save} $r$ for the next iteration.\\
    {\bf Return} $-\frac{1}{\lambda}U D V^T$.
\end{tabbing}
\vspace{10pt}
}
%\noindent\rule[10pt]{.8\textwidth}{0.4pt}

@@

\medskip
\hrule\medskip
\Head{NanoGPT tests}\\
We finetuned NanoGPT by Alex Karpathy on two NVIDIA RTX 4090 GPUs (24 GB each) on tiny stories dataset via 200 iterations. Both Neon and Muon showed better convergence then AdamW. Neon showed similiar convergence rate to Muon

% \begin{figure}[h]
%     \center{\includegraphics[width=0.8\linewidth]{../../figs/mlp24/loss_val.png}}
%     \caption{NanoGPT validation loss}
%     \label{fig:nanogpt_loss}
% \end{figure} 
% \begin{figure}[h]
%     \center{\includegraphics[width=0.8\linewidth]{../../figs/nanogpt/train_loss.pdf}}
%     \caption{NanoGPT train loss}
%     \label{fig:nanogpt_train_loss}
% \end{figure}
\begin{figure}[h]
    \center{\includegraphics[width=0.8\linewidth]{../figs/nanogpt/val_loss.pdf}}
    \caption{NanoGPT validation loss}
    \label{fig:nanogpt_val_loss}
\end{figure}
\medskip
\hrule\medskip
\Head{Conclusion}\\
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
